---
title: "X1,2 follows normal dist Y follows logistic model"
author: "Chenyan Dong 163487177"
date: "17/12/2021"
output: html_document
---

```{r setup, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mvtnorm)
library(tidyverse)
library(bookdown)
library(knitr)
library(rmarkdown)
library(matlib)
```

```{r}
# function for p
expit <- function(p){
  exp(p) / (1 + exp(p))
}
```

The simulated data set contains **2000** observations with variables `id`, `x`, `y`, `Intercept` and `imputed`.\
`X` follows the normal distribution with $\mu = 0$ and $\sigma = 1$.\
`Y` follows the bernoulli distribution, which is a special case of the binomial distribution where a single trial is conducted with $\mu$.\
The $\mu$ could be calculated through the above `expit` function, which follows
$$\textrm{logit} \, P = \textrm{log}(\frac{p}{1 - p}) = \beta_0 + \beta_1 x$$
$$p = \frac{1}{1 + e^{\beta_0 + \beta_1 x}}$$
We set the association between `x` and `y` should be\
$$\textrm{logit} \, \mathbb{E}[\textrm{Y | X} = x]  = \beta_0 + \beta_1 x + \varepsilon$$
where $\beta_0 = -1$ and $\beta_1 = 0.5$ has been set in this simulation.\
We create the variable `Intercept` with 1 for easy to apply the $\beta_0$.\
We randomly select the missing on `x` and replace with **NA**. Then, we create a variable `imputed` that if `x` is **NA** `imputed` would be 1, otherwise would be 0.

```{r base_setup}
# setting
obs <- 2000
id <- rep(1:obs)
X1var <- 1
X2var <- 2
X1X2cov <- -0.25
Yvar <- 1
beta <- c(-1, 0.5, 0.5)
nimpute <- 2

# set seed 
set.seed(163487177)

# x follow the normal distribution with mean = 0, sd = 1
x <- rmvnorm(obs, mean = c(0, 0), sigma = matrix(c(X1var, X1X2cov, X1X2cov, X2var), 2, 2))
x1 <- x[, 1]
x2 <- x[, 2]
# we product a matrix for x 
Xmat <- cbind(1, x1, x2)
rownames(Xmat) <- id
colnames(Xmat) <- c("Intercept", "x1", "x2")

# y follow the bernoulli distribution 
mu <- expit(Xmat %*% beta)
y <- rbinom(obs, 1, mu)

# original dataset
data <- data.frame(id, x1, x2, y)
data$Intercept <- 1

# randomly sample missing on x
miss1 <- with(data, sample(id, 0.05 * obs))
miss2 <- with(data, sample(id, 0.05 * obs))
miss <- unique(c(miss1, miss2))
data[miss1, "x1"] <- NA
data[miss, "x2"] <- NA
data$imputed <- ifelse(data$id %in% miss, 1, 0)
#data
```

We fit a linear regression model of `X` on `Y` is given by $X = \beta_0 + \beta_1 y + \varepsilon$ as the imputation model.

```{r fitting, eval = FALSE}
# fit linear regression model (imputation model)
fit.impute1 <- lm(x1 ~ y, data, y = FALSE, model = FALSE)
fit.impute2 <- lm(x2 ~ y + x1, data, y = FALSE, model = FALSE)
```

We begin to do the imputation on `x` and replace the missing on `x` with imputed values.

```{r imputing, eval = FALSE}
# set seed 
set.seed(163487177)

# impute values X1
variables1 <- c("Intercept", "y")
data_mat1 <- data[, variables1]
desmatrix1 <- rmvnorm(1, mean = fit.impute1$coefficients, sigma = vcov(fit.impute1))
linpred1 <- as.vector(desmatrix1 %*% t(data_mat1))

# account for random noise X1
resid1 <- fit.impute1$residuals
imputed.values1 <- linpred1 + rnorm(length(linpred1), 0, sqrt(X1var))
data.impute1 <- data
data.impute1$x1[miss1] <- imputed.values1[miss]

# impute values X2
variables2 <- c("Intercept", "y", "x1")
data_mat2 <- data.impute1[, variables2]
desmatrix2 <- rmvnorm(1, mean = fit.impute2$coefficients, sigma = vcov(fit.impute2))
linpred2 <- as.vector(desmatrix2 %*% t(data_mat2))

# account for random noise X2
resid2 <- fit.impute2$residuals
imputed.values2 <- linpred2 + rnorm(length(linpred2), 0, sqrt(X2var))
data.impute <- data.impute1
data.impute$x2[miss] <- imputed.values2[miss]

#data.impute
```

**Calculate the imputation model components**

Two of these components $(S^{mis}, d)$ are derived from the imputation model, $f(x_i | \theta)$, where $X$ generally includes multiple variables, some of which are missing for some $i$. Here $f(x_i | \theta)$ represents a joint model for all $X$, which may be specified as a set of conditional models. Values of $(S^{mis}, d)$ are based on the score function and its derivative for the imputation model. 

$$f(x, y | \theta) = f(x | y) f(y)$$
where $f(x | y)$ could be defined as $x = \beta_0 + \beta_1 y + \varepsilon$ as the imputation model is a linear regression and $f(y)$ corresponds to the probability of $Y = 1$ and $Y = 0$. In this case, we do not need to care about the $f(y)$.\

As $\varepsilon \sim N(0, \sigma_\eta^2)$ that $x - \beta_0 - \beta_1 y \sim N(0, \sigma_\eta^2)$.

To evaluate the score function with the log-likelihood function that $s(\theta) = \frac{\partial \, \textrm{log} \, \mathcal{L}(\theta)}{\partial \theta}$.

$$f(x | y) = \frac{1}{\sigma \sqrt{2\pi}} e^{- \frac{1}{2} (\frac{x - \beta_0 - \beta_1 y}{\sigma})^2}$$

$$\textrm{log} \, f = -\frac{1}{2} \textrm{log}(\sigma^2) -\frac{1}{2} (\frac{x - \beta_0 - \beta_1 y}{\sigma})^2$$
$$\frac{\partial \, \textrm{log} \, f}{\partial \beta_0} = - \frac{1}{\sigma^2} \beta_0 + \frac{1}{\sigma^2} (x - \beta_1 y)$$
$$ = \frac{x - \beta_0 - \beta_1 y}{\sigma^2}$$
$$\frac{\partial \, \textrm{log} \, f}{\partial \beta_1} = - \frac{1}{\sigma^2} \beta_1 y^2+ \frac{1}{\sigma^2} (x - \beta_0)y$$
$$ = \frac{(x - \beta_0 - \beta_1 y)y}{\sigma^2}$$
$$\frac{\partial \, \textrm{log} \, f}{\partial \sigma^2} = - \frac{1}{2\sigma^2} + \frac{(x - \beta_0 - \beta_1 y)^2}{2(\sigma^2)^2}$$

```{r score, eval = FALSE}
# evaluate score function X1
imputedvar1 <- data.impute[, "x1"]
imputemod.var1 <- c("Intercept", all.vars(formula(fit.impute1)[-2]))
sigma.est1 <- var(fit.impute1$residuals)
predicted.values1 <- predict(fit.impute1, newdata = data.impute)
S_u_1.1 <- (imputedvar1 - predicted.values1) * data.impute[, imputemod.var1] / sigma.est1
S_sigma_1.1 <- 0.5 * (-1/sigma.est1 + (imputedvar1 - predicted.values1)^2 / sigma.est1^2)
S_u1 <- cbind(S_u_1.1, S_sigma_1.1)
modelvarcount1 <- length(imputemod.var1) + 1
# head(S_u1)

# evaluate score function X2
imputedvar2 <- data.impute[, "x2"]
imputemod.var2 <- c("Intercept", all.vars(formula(fit.impute2)[-2]))
sigma.est2 <- var(fit.impute2$residuals)
predicted.values2 <- predict(fit.impute2, newdata = data.impute)
S_u_1.2 <- (imputedvar2 - predicted.values2) * data.impute[, imputemod.var2] / sigma.est2
S_sigma_1.2 <- 0.5 * (-1/sigma.est2 + (imputedvar2 - predicted.values2)^2 / sigma.est2^2)
S_u2 <- cbind(S_u_1.2, S_sigma_1.2)
modelvarcount2 <- length(imputemod.var2) + 1
# head(S_u2)

# summary together
S_u <- data.frame(S_u1, S_u2)
```

Let $v_i = 1$ denote that the $i^{th}$ observation was observed and $v_i = 0$ denote that the $i^{th}$ observation was imputed. For the $k^{th}$ imputation, $S_k^{mis}$ is a $n \times p$ matrix corresponding to the score of each parameter in the imputation model evaluated for each observation that was imputed. However, if the observation was not imputed, a value of 0 is assigned.

$$S_{i, k}^{mis} = [\frac{\partial \, \textrm{log} \, f(x_i | \theta)}{\partial \theta}|_{\theta = \hat{\theta}}](1 - v_i)$$
$$S_{i, k}^{obs} = [\frac{\partial \, \textrm{log} \, f(x_i | \theta)}{\partial \theta}|_{\theta = \hat{\theta}}]v_i$$

```{r s_mis, eval = FALSE}
# matrices with indicators for whether observation corresponding to each row was missing or not
imputedMat <- matrix((data.impute$id %in% miss), nrow(S_u), ncol(S_u), byrow = FALSE)
observedMat <- 1 - imputedMat

# calculation of S_mis
S_mis <- S_u * imputedMat
```

For calculating the $d$, we need to evaluate the derivative of score function:

$$\frac{\partial^2 \, \textrm{log} \, f}{\partial \beta_0^2} = - \frac{1}{\sigma^2}$$

$$\frac{\partial^2 \, \textrm{log} \, f}{\partial \beta_1^2} = - \frac{y^2}{\sigma^2}$$

$$\frac{\partial^2 \, \textrm{log} \, f}{\partial \beta_0 \, \partial \sigma^2} = - \frac{x - \beta_0 - \beta_1 y}{(\sigma^2)^2}$$
$$\frac{\partial^2 \, \textrm{log} \, f}{\partial \beta_1 \, \partial \sigma^2} = - \frac{(x - \beta_0 - \beta_1 y)y}{(\sigma^2)^2}$$
$$\frac{\partial^2 \, \textrm{log} \, f}{\partial (\sigma^2)^2} = \frac{1}{\sigma^2} + \frac{1}{2}(-2 \frac{x - \beta_0 - \beta_1 y}{(\sigma^2)^3})$$
$$ = \frac{1}{\sigma^2} - \frac{x - \beta_0 - \beta_1 y}{(\sigma^2)^3}$$

```{r score_der, eval = FALSE}
# evaluate derivative of score function
imputedvar1 <- data.impute[, "x1"]
imputemod.var1 <- c("Intercept", all.vars(formula(fit.impute1)[-2]))
impvar_n1 <- length(imputemod.var1)
sigma.est1 <- var(fit.impute1$residuals)
predicted.values1 <- predict(fit.impute1, newdata = data.impute)

S2_uu1 <- matrix(NA, nrow(data.impute), length(imputemod.var1))
S2_uu1 <- with(data.impute, 
              (-1/sigma.est1 * t(data.impute[-miss1, imputemod.var1]) %*% 
                 t(t(data.impute[-miss1, imputemod.var1]))))
S2_usigma1 <- with(data.impute, 
                  (-1/sigma.est1^2 * t(data.impute[-miss1, c(imputemod.var1)]) %*%
                     (imputedvar1 - predicted.values1)[-miss1]))
S2_sigmasigma1 <- sum(with(data.impute, ifelse(data.impute$imputed == 1, 0, 
                                              1/(2*sigma.est1^2) - (imputedvar1 - predicted.values1)^2/sigma.est1^3)))

S2_mod1 <- matrix(0, impvar_n1 + 1, impvar_n1 + 1)
S2_mod1[1:impvar_n1, 1:impvar_n1] <- S2_uu1/dim(data.impute)[1]
S2_mod1[1:impvar_n1, impvar_n1 + 1] <- S2_mod1[impvar_n1 + 1, 1:impvar_n1] <- apply(S2_usigma1, 1, mean)/dim(data.impute)[1]
S2_mod1[impvar_n1 + 1, impvar_n1 + 1] <- S2_sigmasigma1/dim(data.impute)[1]
#S2_mod1

# evaluate derivative of score function
imputedvar2 <- data.impute[, "x2"]
imputemod.var2 <- c("Intercept", all.vars(formula(fit.impute2)[-2]))
impvar_n2 <- length(imputemod.var2)
sigma.est2 <- var(fit.impute2$residuals)
predicted.values2 <- predict(fit.impute2, newdata = data.impute)

S2_uu2 <- matrix(NA, nrow(data.impute), length(imputemod.var2))
S2_uu2 <- with(data.impute, 
              (-1/sigma.est2 * t(data.impute[-miss, imputemod.var2]) %*% 
                 t(t(data.impute[-miss, imputemod.var2]))))
S2_usigma2 <- with(data.impute, 
                  (-1/sigma.est2^2 * t(data.impute[-miss, c(imputemod.var2)]) %*%
                     (imputedvar2 - predicted.values2)[-miss]))
S2_sigmasigma2 <- sum(with(data.impute, ifelse(data.impute$imputed == 1, 0, 
                                              1/(2*sigma.est2^2) - (imputedvar2 - predicted.values2)^2/sigma.est2^3)))

S2_mod2 <- matrix(0, impvar_n2 + 1, impvar_n2 + 1)
S2_mod2[1:impvar_n2, 1:impvar_n2] <- S2_uu2/dim(data.impute)[1]
S2_mod2[1:impvar_n2, impvar_n2 + 1] <- S2_mod2[impvar_n2 + 1, 1:impvar_n2] <- apply(S2_usigma2, 1, mean)/dim(data.impute)[1]
S2_mod2[impvar_n2 + 1, impvar_n2 + 1] <- S2_sigmasigma2/dim(data.impute)[1]
#S2_mod2
```

To calculate $d$, we first take derivates of the score function with respect to each parameter, evaluate at each observation that was not imputed, and take the average. We then take the inverse of this $p \times p$ matrix, multiply by the transpose of $S^{obs}$, and multiply by -1. Note that $S^{obs}$ is the score evaluated for each observation that was not imputed; imputed values are assigned a value of 0.

$$d_i^T = -[\frac{1}{n} \sum^{n}_{i = 1} v_i \, \frac{\partial}{\partial \theta^T}(\frac{\partial \, \textrm{log} \, f(x_i | \theta)}{\partial \theta})|_{\theta = \hat{\theta}}]^{-1} \, {S_i^{obs}}^T$$

```{r d, eval = FALSE}
# calculation of D
S_obs <- S_u * observedMat
n_S2mod1 <- ncol(S2_mod1)
n_S2mod2 <- ncol(S2_mod2)
n_S2mod <- n_S2mod1 + n_S2mod2
	
S2 <- matrix(0, n_S2mod, n_S2mod)
S2[1:n_S2mod1, 1:n_S2mod1] <- S2_mod1
S2[(n_S2mod1 + 1):n_S2mod, (n_S2mod1 + 1):n_S2mod] <- S2_mod2

Dmat <- solve(S2)
d_t <- ((-1)*Dmat) %*% t(S_obs)
d <- t(d_t)
```

We fit a logistic regression model of `Y` on `X` is given by $\textrm{logit} \, \mathrm{E}(\textrm{Y | X} = x)  = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \varepsilon$ as the analysis model.

```{r, analysis, eval = FALSE}
# fit logistic regression model (analysis model)
fit.analysis <- glm(y ~ x1 + x2, data.impute, family = "binomial")

# set for analysis mode and data
analysis.predict <- predict(fit.analysis, type = "response")
analysis.vars <- names(fit.analysis$coef)
analysis.vars <- ifelse(analysis.vars == "(Intercept)", "Intercept", analysis.vars)

# capture estimates and corresponding SE from each imputation
# analysis.est <- fit.analysis$coef["x"]
# analysis.se <- sqrt(diag(vcov(fit.analysis))["x"])
```

<p>&nbsp;</p>

**Analysis model components**

Both $(u, \tau)$ components are based on the estimating equations pertaining to the analysis model. For each imputation, we evaluate the estimating equation $u_i(\hat{\theta}, \hat{\beta}_k)$ for all subjects in the analysis dataset.

$$u_i(\hat{\theta}, \hat{\beta}_k) = (y - \beta_0 - \beta_1 x)y$$

```{r u, eval = FALSE}
# calculation of u
# evaluate the estimating equation for each observation
U_i <- data.impute[, analysis.vars] * matrix((data.impute$y - analysis.predict), dim(data.impute[, analysis.vars]))
U_i[is.na(U_i)] <- 0	
```

To calculate $\tau$ , we take the derivative of the estimating equation and evaluate as:

$$\tau = -\frac{1}{nm} \sum_{i = 1}^n \sum_{k = 1}^m (\frac{\partial \, u(i, k)}{\partial \beta^T})$$

```{r tau, eval = FALSE}
# calculation of tau
# take the derivative of the estimating equation and evaluate for each observation
tempdat <- data.impute[, analysis.vars]
#predicted.values is actually predicted LP
analysis.predLP <- predict(fit.analysis, newdata = tempdat, type = "link")
tau_i <- matrix(NA, ncol(tempdat), ncol(tempdat))
rownames(tau_i) <- colnames(tau_i) <- colnames(tempdat)
constant.term <- exp(analysis.predLP)/(1 + exp(analysis.predLP))^2
for (matrow in 1:ncol(tempdat)){
  for (matcol in 1:ncol(tempdat)){
    temp <- (-1)*(tempdat[, matrow]) * ((tempdat[, matcol])) * constant.term
    tau_i[matrow, matcol] <- (-1)*sum(temp)
	}
}
```

```{r imputation, warning = FALSE}
imputation <- function(data, nimpute, obs) {
  # initialize select variables
  results <- replicate(800, {
    set.seed(163487177)
  U_i_sum <- 0
  kappa_sum <- 0
  tau_sum <- 0
  analysis.est1 <- analysis.se1 <- analysis.est2 <- analysis.se2 <- NULL
  
  for(p in 1:nimpute){
    <<fitting>>
    <<imputing>>
    <<score>>
    <<score_der>>
    <<s_mis>>
    <<d>>
    <<analysis>>
    # capture estimates and corresponding SE from each imputation
    analysis.est1[p] <- fit.analysis$coef["x1"]
    analysis.est2[p] <- fit.analysis$coef["x2"]
    analysis.se1[p] <- sqrt(diag(vcov(fit.analysis))["x1"])
    analysis.se2[p] <- sqrt(diag(vcov(fit.analysis))["x2"])
    <<u>>
    <<tau>>
    # create summations of certain components across imputations
    U_i_sum <- U_i_sum + U_i	
	  tau_sum <- tau_sum + tau_i
	  # calculate kappa for a given imputation
	  kappa <- t(U_i) %*% t(t(S_mis))
	  # create summation for kappa across imputations
	  kappa_sum <- kappa_sum + kappa
  }
  u_bar <- U_i_sum/nimpute
	u_bar <- t(t(u_bar))

	omega <- (t(u_bar) %*% t(t(u_bar)))/(obs)
	kappa <- t(t(kappa_sum/(obs * nimpute)))

	alpha <- (t(d) %*% d) / obs

	delta <- omega + kappa %*% alpha %*% t(kappa) + (1/obs) * (kappa %*% t(d) %*% u_bar + t(kappa %*% t(d) %*% u_bar))
	tau <- tau_sum/(nimpute * obs)
	gamma <- (1/obs) * t(t(solve(tau))) %*% delta %*% t(solve(tau)) 
	
	#Save SE estimates to calculate CIs!
	seRW <- sqrt(diag(gamma)[c("x1", "x2")])
	seRR <- sqrt(c(mean(analysis.se1^2), mean(analysis.se2^2)) + (nimpute + 1)/nimpute * c(var(analysis.est1), var(analysis.est1)))
	
	# parameter Estimate
	estimate1 <- mean(analysis.est1)
	estimate2 <- mean(analysis.est2)
	estimate <- c(estimate1, estimate2)
	
	c(estimate, seRW, seRR)
  })
  
  estimate1 <- mean(results[1,])
  estimate2 <- mean(results[2,])
  seRW1 <- mean(results[3,])
  seRW2 <- mean(results[4,])
  seRR1 <- mean(results[5,])
  seRR2 <- mean(results[6,])
  
  upperCI_RW1 <- estimate1 + 1.96 * seRW1
  upperCI_RR1 <- estimate1 + 1.96 * seRR1
  lowerCI_RW1 <- estimate1 - 1.96 * seRW1
  lowerCI_RR1 <- estimate1 - 1.96 * seRR1
  
  upperCI_RW2 <- estimate2 + 1.96 * seRW2
  upperCI_RR2 <- estimate2 + 1.96 * seRR2
  lowerCI_RW2 <- estimate2 - 1.96 * seRW2
  lowerCI_RR2 <- estimate2 - 1.96 * seRR2
  
  
  output <- rbind(c(estimate1, seRW1, lowerCI_RW1, upperCI_RW1, seRR1, lowerCI_RR1, upperCI_RR1), 
                  c(estimate2, seRW2, lowerCI_RW2, upperCI_RW2, seRR2, lowerCI_RR2, upperCI_RR2))
	rownames(output) <- c("x1", "x2")
	colnames(output) <- c("estimate", "seRobinsWang", "95%RWl", "95%RWu", "seRubin", "95%RRl", "95%RRu")
	output
}
```



```{r, warning = FALSE}
imputation(data, nimpute, obs)
```

